Dependent Object Types

### 2012-09-04 doc/candidacy/report.tex
typos

It is straightforward to encode this example in Scala, with the
correspondence Object = Module, and Object Type = Signature. Indeed,
the ${\nu}Obj$ calculus, which we study in section~\ref{sec:nuObj}
embraces this correspondence, and futher claims: Class = Method =
Functor, as we will see. Note that type members, and hence
path-dependent types, are crucial for this correspondence.
\lstinputlisting{ml.scala}

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allow {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with its methods and fields. {\it
  Further binding} enables the nested classes to be further
extended. Path-dependent types arise naturally to distinguish objects
coming from different families. Yet, thanks to {\it family
  polymorphism}, code written for one family also works for extension
of that family. In Tribe, path types can depend simultaneously on both
classes and objects. Additionally, paths can use an \code{out} field
to move from an object to the object which surrounds it. This
\code{out} feature enables ubiquitous access to an object's family
without the need to drag around family arguments.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and \code{out} selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColourGraph} are no longer in a
subclassing / further-binding with one another, as this would only bring
confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
\code{out} faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the Tribe paper, using \code{out} in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

A class template $\cls{x: S \sep \tup d}$ is closely related to a
translucent sum of section~\ref{sec:homl}, except that it needs to be instantiated into an
object while a translucent sum is a module. Like in a translucent sum,
$\tup d$ binds term labels to values and type labels to types. The $x$
of type $S$ stands for the self, i.e.\ the object being constructed
from the template. Its scope is the bindings $\tup d$. Since the self
is explicitly typed, the bindings in class templates can
``contractively'' refer to each other (and even to themselves) via
$x$, while recall that, in translucent sums, only later fields can
refer to earlier ones but the references don't have any
``contractive'' restrictions. Roughly, ``contractiveness'' in
${\nu}Obj$ guarantees that a field cannot be referenced before it is
initalized.

@@

A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
class with an abstract member \code!arg! for the function argument:
\begin{lstlisting}
[x: {arg: T}| fun = [res = $t'$]]
\end{lstlisting}
$t'$ corresponds to term $t$ in which all occurrences of $x$ are
replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
the right-hand side of \code!fun!: this wouldn't be ``contractive'',
because \code!arg! might not be initialised before \code!fun!. Hence,
we pack $t'$ in another class to ensure contractiveness. Now, an
application $g(e)$ gets decomposed into three subsequent steps:
\begin{lstlisting}
$\nu\,$g$_{app}$  <- g & [arg = e];
$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
g$_{eval}$.res
\end{lstlisting}
First we instantiate function \code!g! with a concrete argument
yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
creating an instance \code!g$_{eval}$! of it. Finally we extract the
result by querying field \code!res! of \code!g$_{eval}$!.

>>>>

It is straightforward to encode this example in Scala, with the
correspondence Object = Module, and Object Type = Signature. Indeed,
the ${\nu}Obj$ calculus, which we study in section~\ref{sec:nuObj}
embraces this correspondence, and further claims: Class = Method =
Functor, as we will see. Note that type members, and hence
path-dependent types, are crucial for this correspondence.
\lstinputlisting{ml.scala}

@@

In~\cite{tribe}, Clarke et al. present a calculus for virtual
classes. Virtual classes allow {\it families} of classes to be
inherited, rather than just single classes. Classes are lexically
nested inside other classes. When a class is inherited, its nested
inner classes are inherited along with their methods and fields. {\it
  Further binding} enables the nested classes to be further
extended. Path-dependent types arise naturally to distinguish objects
coming from different families. Yet, thanks to {\it family
  polymorphism}, code written for one family also works for extension
of that family. In Tribe, path types can depend simultaneously on both
classes and objects. Additionally, paths can use an \code{out} field
to move from an object to the object which surrounds it. This
\code{out} feature enables ubiquitous access to an object's family
without the need to drag around family arguments.

@@

Now, we can emulate this Tribe example in Scala at the expense of some
syntactic overhead, since further binding and \code{out} selection are
not native. We use virtual types to ensure that the inner classes are
indeed ``virtual'', i.e. dynamically instead of statically
dispatched. In order to conform to the ``virtual class'' pattern, the
leaf classes \code{Graph} and \code{ColouredGraph} are no longer in a
subclassing / further-binding with one another, as this would only bring
confusion when ascribing a coloured graph to a graph, since the constructors are statically dispatched. Emulating
\code{out} faithfully involves a bit of hackery in making Scala recognize
that, for example, \code{e.from.type} is a subtype of
\code{e.out.Node}. As noted in the Tribe paper, using \code{out} in library
code obscures the symmetry between the parameters (for example, in the
\code{distance} method below) but allows the family object to be
recovered without being explicitly passed around (contrast with the
method \code{distance2}).
\lstinputlisting{tribe.scala}

@@

A class template $\cls{x: S \sep \tup d}$ is closely related to a
translucent sum of section~\ref{sec:homl}, except that it needs to be instantiated into an
object while a translucent sum is a module. Like in a translucent sum,
$\tup d$ binds term labels to values and type labels to types. The $x$
of type $S$ stands for the self, i.e.\ the object being constructed
from the template. Its scope is the bindings $\tup d$. Since the self
is explicitly typed, the bindings in class templates can
``contractively'' refer to each other (and even to themselves) via
$x$, while recall that, in translucent sums, only later fields can
refer to earlier ones but the references don't have any
``contractive'' restrictions. Roughly, ``contractiveness'' in
${\nu}Obj$ guarantees that a field cannot be referenced before it is
initialized.

@@

A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
class with an abstract member \code!arg! for the function argument:
\begin{lstlisting}
[x: {arg: T}| fun = [res = $t'$]]
\end{lstlisting}
$t'$ corresponds to term $t$ in which all occurrences of $x$ are
replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
the right-hand side of \code!fun!: this wouldn't be ``contractive'',
because \code!arg! might not be initialized before \code!fun!. Hence,
we pack $t'$ in another class to ensure contractiveness. Now, an
application $g(e)$ gets decomposed into three subsequent steps:
\begin{lstlisting}
$\nu\,$g$_{app}$  <- g & [arg = e];
$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
g$_{eval}$.res
\end{lstlisting}
First we instantiate function \code!g! with a concrete argument
yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
creating an instance \code!g$_{eval}$! of it. Finally we extract the
result by querying field \code!res! of \code!g$_{eval}$!.
